DCMIP2016, Part 1: Models and Equation Sets

### 2016-08-18 contributions/FVM/FVM-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

@@

\Author[6]{Thomas}{Dubos}
\Author[7]{Marco}{Giorgetta}
\Author[8]{Elijah}{Goodfriend}
\Author[9]{David A.}{Hall}
\Author[10]{Lucas}{Harris}
\Author[8]{Hans}{Johansen}
\Author[11]{Christian}{Kuehnlein}
\Author[12]{Vivian}{Lee}
\Author[13]{Thomas}{Melvin}
\Author[14]{Hiroaki}{Miura}
\Author[15]{David}{Randall}
\Author[16]{Alex}{Reinecke}
\Author[4]{William}{Skamarock}
\Author[16]{Kevin}{Viner}
\Author[17]{Robert}{Walko}

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{Institut Pierre-Simon Laplace (IPSL)}
\affil[7]{Max Planck Institute for Meteorology}
\affil[8]{Lawrence Berkeley National Laboratory}
\affil[9]{University of Colorado, Boulder}
\affil[10]{Geophysical Fluid Dynamics Laboratory}
\affil[11]{European Center for Medium-Range Weather Forecasting}
\affil[12]{Environment Canada}
\affil[13]{U.K. Met Office}
\affil[14]{University of Tokyo}
\affil[15]{Colorado State University}
\affil[16]{Naval Research Laboratory}
\affil[17]{University of Miami}

@@

@@

\begin{table}[h]
\caption{List of symbols used in this manuscript} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$\delta$ & Divergence\\
$\zeta$ & Relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\Theta$ & Potential temperature \\
$\Theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

%%%GFDL start

FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}. The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm. The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side. 
%%%GFDL end

@@

@@

\subsection{FV$^3$} \label{sec:FV3Equations}

%%%GFDL start
FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
%%%GFDL end

%They can be written as:%
%
%\begin{align}
%\pdiff{\rho}{t} &= - \nabla \cdot (\rho \vb{u}), \\
%\pdiff{\vb{u}}{t} &= \left( f\vg{\hat k + \zeta} \right ) \times \vb{u} - \frac{1}{\rho}\nabla p |_z, \\
%\pdiff{\theta_v}{t} &= - \vb{u} \cdot \nabla \theta_v, \\
%\pdiff{w}{t} + \vb{u} \cdot \nabla w &= -\frac{1}{\rho}\pdiff{p}{z} - g,
%\end{align} in terms of Kinetic energy $K = \vb{u} \cdot \vb{u}$, relative vorticity $\vg{\zeta} = \nabla \times \vb{u}$, and local Coriolis parameter $f$. Here, gradient operators (except for that in the pressure-gradient term) are along Lagrangian surfaces.

@@

{\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
%%%GFDL start

Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations. The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.  FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion. The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow. If additional damping is desired, or if the monotonicity constraint has been disabled, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion. This dissipation is of the same order of accuracy as the divergence damping; however the nondimensional coefficient needed is much smaller, usually by at least 3--4 times, than the divergence damping. Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep; there is no explicit damping perpendicular to the Lagrangian surfaces. An option to convert kinetic energy lost to the hyperdiffusion to heat is available. FV$^3$ can also apply a horizontal Smagorinsky eddy diffusion if desired.

A substantial amount of wave absorption is provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain. An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.

%%%GFDL end

@@

\subsection{Energy fixers}
%%%GFDL start

FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment. This is only done before the physics is called and is not used in idealized simulations. 
%%%GFDL end

@@

\subsection{Forward-backward vertically-Lagrangian dynamics}

%%%GFDL start

FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics. With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}. The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the nonhydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver. This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps. Physics tendencies are applied ``impulsively'' at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep. 
%%%GFDL end

@@

\subsection{Semi-Implicit time integration}

@@

\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV$^3$)}
%%%GFDL start

FV$^3$ uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate. The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms. Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass. The nonhydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.

%%%GFDL end

@@

\subsection{ECMWF PentaRei Finite Volume Model}

{\color{red}[KUEHNLEIN]}

@@

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{DCMIP2016.bib,FV3.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

>>>>

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
\usepackage{subcaption}
%\usepackage{subfig}
%\usepackage{rotating}

@@

\Author[6]{Thomas}{Dubos}
\Author[7]{Marco}{Giorgetta}
\Author[8]{Elijah}{Goodfriend}
\Author[9]{David A.}{Hall}
\Author[10]{Lucas}{Harris}
\Author[8]{Hans}{Johansen}
\Author[11]{Christian}{K\"uhnlein}
\Author[12]{Vivian}{Lee}
\Author[13]{Thomas}{Melvin}
\Author[14]{Hiroaki}{Miura}
\Author[15]{David}{Randall}
\Author[16]{Alex}{Reinecke}
\Author[4]{William}{Skamarock}
\Author[16]{Kevin}{Viner}
\Author[17]{Robert}{Walko}

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{Institut Pierre-Simon Laplace (IPSL)}
\affil[7]{Max Planck Institute for Meteorology}
\affil[8]{Lawrence Berkeley National Laboratory}
\affil[9]{University of Colorado, Boulder}
\affil[10]{Geophysical Fluid Dynamics Laboratory}
\affil[11]{European Centre for Medium-Range Weather Forecasting}
\affil[12]{Environment Canada}
\affil[13]{U.K. Met Office}
\affil[14]{University of Tokyo}
\affil[15]{Colorado State University}
\affil[16]{Naval Research Laboratory}
\affil[17]{University of Miami}

@@

% User-defined mathematical symbols
\renewcommand{\emph}[1]{{\color{red}\textbf{#1}}}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\p}[1]{\partial #1}
%Then, writing a partial is as simple as $\pd{J}{x}$.
\newcommand{\ol}[1]{{\overline #1}}
\newcommand{\mbf}[1]{{\mathbf #1}}
\newcommand{\mbfol}[1]{\overline{{\mathbf #1}}}
\newcommand{\wdh}[1]{{\widehat #1}}
\newcommand{\wdt}[1]{{\widetilde #1}}
\newcommand{\sss}[1]{{\scriptscriptstyle #1}}
\newcommand{\scs}[1]{{\scriptstyle #1}}

@@

\begin{table}[h]
\caption{List of symbols used in this manuscript} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Pressure velocity  \\
$\delta$ & Divergence\\
$\zeta$ & Relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q_v$ & Water vapor water mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

@@

\subsection{Octahedral reduced Gaussian grid}
\label{subsection:octahedralreducedgaussiangrid}

As with the classical reduced Gaussian grid of \cite{hortalsimmonsMWR1991}, the octahedral 
reduced Gaussian grid \citep{malardel2016,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
roots of the Legendre polynomials. 
The two grids differ in the arrangement of the points along the latitudes, which follows a simple
rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
are added with every latitude towards the equator, whereby the spacing between points along the 
latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
is suitable for transformations involving spherical harmonics, and has been introduced for global 
medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).

\begin{figure}
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=5cm]{octahedral_gauss_nodes}
 \end{subfigure}%
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=4.7cm]{octahedral_RGG_white}
 \end{subfigure}
\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
\label{fig:octahedralgrid}
\end{figure}

%octahedral_RGG_white.pdf

%Fig.~\ref{fig:octahedralgridFVM}

@@

\subsection{FVM} \label{sec:FVMEquations}

The fully compressible Euler equations solved in FVM are given as
\begin{subequations} \label{fvm:compressible}
\begin{align}
 & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
\label{fvm:mass}  \\
& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\epsilon q'_v - q_c - q_r) \right)
- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
+ \mathbf{M} \right)~,
\label{fvm:momentum} \\
%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
+  \mathcal{H} \right)~,
\label{fvm:thermodynamic} \\
& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
\label{fvm:gaslaw}
\end{align}
\end{subequations}
which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
and dry entropy \eqref{fvm:thermodynamic}.
Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
pressure perturbation $\pi'$, 
with the thermodynamic variables related by the gas law 
\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
respectively. All primed variables correspond to deviations from an ambient state (denoted by
subscript "a")  that satisfies 
a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
see \cite{prusa2008} and \cite{smolarkiewiczJCP2014}. 
The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
coefficient 
\begin{equation}
\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
\end{equation}
in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
$\epsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
and the transformation of the physical to the contravariant velocity 
$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003} and \cite{kuehnleinJCP2012} for discussion.
%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
in \eqref{fvm:thermodynamic} represents the diffusion of heat.

%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.

@@

{\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}

@@

\subsection{Energy fixers}

@@

@@

\subsection{Semi-implicit time integration of the fully compressible Euler equations in FVM}

In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
moist dynamics. 
The generic two-time-level second-order template algorithm employed in the integration is given as
\begin{equation}
\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
+0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
\label{fvm:solscalar}
\end{equation}
In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}~\footnote{Furthermore, 
in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.}.

The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
momentum \eqref{fvm:momentum}, and moisture equations enter
\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term. 
The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
executed in an outer iteration.

%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
%slow rotational modes.

%The integration scheme of FVM approximates the generic conservation law
%\begin{equation}
%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
%\label{scalar}
%\end{equation}
%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
%second-order template algorithm
%\begin{equation}
%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
%+0.5\,\delta t\,R^{\Psi}|_{\mbf{i}}^{\scs n+1}~,
%\hspace{1em} \widetilde{\Psi}^{n}\,\equiv\,\Psi^n+0.5\,\delta t\,R^{\Psi}|^{\scs n}~,
%%\hspace{1em} \widetilde{\psi}^{n}$\,$\equiv$\,$\psi^n+0.5\,\delta\ol{t}\,R^{\psi}|^{\scs n}~,
%\label{solscalar}
%\end{equation}
%where $\mathcal{A}$ is a shorthand for the advective transport operator MPDATA. The index $\mathbf{i}$ 
%denotes the spatial position on the computational grid, $\delta t$ is the time step size between levels $n$
%and $n+1$,

@@

\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV3)}

{\color{red}[HARRIS]}

@@

\subsection{Finite-volume module of the Integrated Forecasting System}
\label{sec:FVM}

%{\color{red}[KUEHNLEIN]}

The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is developed at 
ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
are available by means of simple switches. The formulation incorporates 
a generalised, optionally time-dependent, terrain-following vertical coordinate based on height. 
A centred two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
Conjugate Residual approach. 
The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
The horizontal spatial discretisation is fully unstructured finite-volume using the 
median-dual approach. This is combined with a structured-grid finite-difference approach in the 
vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
In both the horizontal and the vertical discretisation, all prognostic variables are 
co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.

No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
boundary layer parametrisation and the constant-coefficient second-order dissipation/diffusion in 
the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
at every time step according to a given maximum advective CFL number (typically somewhat smaller 
than 1). The physics time step is identical to the dynamics time step. 
%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.

%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
%efficiency and accuracy.

%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
%framework developed at ECMWF.

@@

%% Since the Copernicus LaTeX package includes the TeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{DCMIP2016,../../bib/fvm}
%\bibliography{fvm}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}
